Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
describe nonreflective syntax

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App:
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
          {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
          {\filterset{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \x{} is a \Number{}.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\x{}}{\Number}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
$$
By T-App,
\inc{} requires a \Number{} argument which we just proved.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
Finally we conclude with T-Do using the conditional and increment expressions as premises.
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
          {\hastype{\doexp
                     {\ifexp{\appexp{\numberhuh{}}{\x{}}}
                        {\nil{}}
                        {\errorvalv{}}}
                     {\appexp{\inc{}}{\x{}}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
type hints to help the compiler eliminate reflective calls.
In this section we introduce our model of Java, including syntax
for unresolved Java

@@

We extend the external language with Java field lookups and calls to
methods and constructors.
These come without typing or evaluation rules:
Typed Clojure requires the Clojure compiler to statically resolve all Java interactions
in typed code.
Type hinted expressions such as \localhintsyntax{} 
to help the compiler eliminate reflection.

@@

$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{} and
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.

%$$
%\TALetHint{}
%$$

We now have two languages: an internal language that may contain non-reflective Java calls (figure~\ref{main:figure:javatyping}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).

>>>>

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App:
$$
\judgement{\isprop{\UnionNilNum}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\Nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\UnionNilNum}{\x{}}}
          {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
          {\filterset{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \x{} is a \Number{}.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\x{}}{\Number}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
$$
By T-App,
\inc{} requires a \Number{} argument which we just proved.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
Finally we conclude with T-Do using the conditional and increment expressions as premises.
$$
\judgement{\isprop{\UnionNilNum}{\x{}}}
          {\hastype{\doexp
                     {\ifexp{\appexp{\numberhuh{}}{\x{}}}
                        {\nil{}}
                        {\errorvalv{}}}
                     {\appexp{\inc{}}{\x{}}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
type hints to help the compiler eliminate reflective calls.
In this section, first we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this we add notation for
\emph{non-reflective} Java forms that carry sufficient tag information 
to unambiguously identify a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.

@@

We extend the syntax with Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form 
These come without typing or evaluation rules---Typed Clojure 
requires the Clojure compiler to statically resolve all Java interactions
in typed code, so they are implicitly type errors
(as demonstrated in section~\ref{sec:overviewjavainterop}).
We add type hinted syntax for local variables and let-bindings.

Now we model the compiler's reflection-resolution algorithm.
Figure~\ref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
$$
{\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});
$$
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with signature
$$
{\classhint{2}}\ \fld{};
$$
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor
$$
{\classhint{}}\ ({\overrightarrow{\classhint{i}}});
$$

@@

$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{} and
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$
